Global well-being and mental health in the internet age

### 2023-01-30 ms.Rmd
Updates

---
title: Global well-being and mental health in the internet age
shorttitle: Global well-being
leftheader: Global well-being
author: 
  - name: Matti Vuorre
    affiliation: 1
    address: Oxford Internet Institute, University of Oxford
    email: matti.vuorre@oii.ox.ac.uk
  - name: Andrew K. Przybylski
    affiliation: 1
    corresponding: yes
    address: Oxford Internet Institute, University of Oxford
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Oxford Internet Institute, University of Oxford
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.}
abstract: | 
  In the last two decades the rise of internet technologies have impacted nearly all domains of human life and inspired concern that their widespread adoption has impacted mental health and psychological well-being. However, research on the topic is contested and hampered by methodological shortcomings leaving the broader consequences of internet adoption unknown. We show that the past two decades have seen small and inconsistent changes in global well-being and mental health. The broadest available health and internet technology data does not support the idea that the adoption of internet and mobile broadband is consistently linked to negative psychological outcomes. However, associations between internet technology adoption and depression and self-harm were strongest for individuals aged between 10 and 34. Further investigation of this topic requires transparent study of online behaviours where they occur, namely on online platforms. We call for increased collaborative efforts between independent scientists and the internet technology sector.

@@

We analysed the data with meta-analytic bayesian hierarchical regression models [@burknerBrmsPackageBayesian2017; @gelmanDataAnalysisUsing2007; @standevelopmentteamStanModelingLanguage2021]. We first asked how the well-being outcomes had changed over time. To answer, we modelled each outcome (e.g. life satisfaction) on population level intercepts (we centred time on 2010), contrasts of time (in continuous years), sex (Female: 1, Male: -1), and their interaction. We allowed all coefficients to vary across countries, age groups, and the age by country interactions, and treated all outcomes as normally distributed.

@@

Our third research question asked whether any trends in well-being or its associations with internet or mobile broadband adoption were specific to adolescents or other demographic groups. To answer, the model allowed all parameters to vary randomly across the age groups and the age by country interaction. Therefore, each age group, on average and within each country, received their own partially pooled estimates [@gelmanDataAnalysisUsing2007]. This bayesian approach to estimating age-specific associations is beneficial, given that the large number of age groups, especially within countries, would otherwise present difficulties with uncertain estimates and multiple comparisons. Therefore, we could compare younger age groups to older age groups, and investigate other contrasts, with confidence and without additional post-hoc adjustment procedures [@gelmanWhyWeUsually2012].

@@

@@

Our model also measured the extent to which internet technology adoption predicted well-being between countries. These between-country associations indicated, with high confidence, that countries with greater average levels of internet and mobile broadband adoption tended to report greater average levels of life satisfaction and positive experiences, and lower average levels of negative experiences (Figure \@ref(fig:fig-gwp-data)B, Between-country associations). However, this pattern is likely to reflect multiple other causes impinging on both well-being and internet technology adoption.

@@

@@

Our results do not provide evidence supporting the view that the internet, and technologies enabled by it such as smartphones with internet access, are actively promoting or harming either well-being or mental health globally. Our results also suggest that age- and sex-based differences in associations linking internet technology adoption and well-being are small. However, we did find evidence suggesting that internet technology adoption is more negatively associated with young individuals' mental health than older groups' (whose associations were, in some cases, positive).

It is important to note that the data and theory required to address this question at the causal level are absent. Consequently, our analyses cannot account for potential confounders in the associations linking mental health and well-being to internet technology adoption. Our descriptions, therefore, are suggestive but are not intended to provide evidence for, or against, causal relations. Although the first to study this topic within a global perspective, future studies should investigate other potential differences based on demographic, economic, and cultural factors, which we could not address here. In addition, our results regarding mental health, while comprehensive, are necessarily and considerably less certain than those on well-being, due to the lack of diverse global data and our own reliance on GBD estimates in place of objective measurements of mental health.

>>>>

---
title: Global well-being and mental health in the internet age
shorttitle: Global well-being
leftheader: Global well-being
author: 
  - name: Matti Vuorre
    affiliation: 1
    address: Oxford Internet Institute, University of Oxford
    email: matti.vuorre@oii.ox.ac.uk
  - name: Andrew K. Przybylski
    affiliation: 1
    corresponding: yes
    address: Oxford Internet Institute, University of Oxford
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Oxford Internet Institute, University of Oxford
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.}
abstract: | 
  In the last two decades the rise of internet technologies have impacted nearly all domains of human life and inspired concern that their widespread adoption has impacted mental health and psychological well-being. However, research on the topic is contested and hampered by methodological shortcomings leaving the broader consequences of internet adoption unknown. We show that the past two decades have seen small and inconsistent changes in global well-being and mental health. The broadest available health and internet technology data does not support the idea that the adoption of internet and mobile broadband is consistently linked to negative psychological outcomes. However, associations between internet technology adoption and depression and self-harm were strongest (albeit still small) for individuals aged between 10 and 34. Further investigation of this topic requires transparent study of online behaviours where they occur, namely on online platforms. We call for increased collaborative efforts between independent scientists and the internet technology sector.

@@

We analysed the data with meta-analytic bayesian hierarchical regression models [@burknerBrmsPackageBayesian2017; @gelmanDataAnalysisUsing2007; @standevelopmentteamStanModelingLanguage2021]. We first asked how the well-being outcomes had changed over time. Bayesian methods are especially suitable to our reserch questions, because they involve many potential contrasts between e.g. countries and ages. Using bayesian hierarchical models allowed us to accurately quantify uncertainty at different levels of analyses while avoiding problems with multiple comparisons (see below). To answer, we modelled each outcome (e.g. life satisfaction) on population level intercepts (we centred time on 2010), contrasts of time (in continuous years), sex (Female: 1, Male: -1), and their interaction. We allowed all coefficients to vary across countries, age groups, and the age by country interactions, and treated all outcomes as normally distributed.

@@

Our third research question asked whether any trends in well-being or its associations with internet or mobile broadband adoption were specific to adolescents or other demographic groups. To answer, the model allowed all parameters to vary randomly across the age groups and the age by country interaction. Therefore, each age group, on average and within each country, received their own partially pooled estimates [@gelmanDataAnalysisUsing2007]. This bayesian approach to estimating age-specific associations is beneficial, given that the large number of age groups, especially within countries, would otherwise present difficulties with uncertain estimates and multiple comparisons. In addition, bayesian methods allowed quantifying uncertainties at different levels of analysis, rather than providing e.g. only point estimates of country-specific quantities. Therefore, we could compare younger age groups to older age groups, and investigate other contrasts, with confidence and without additional post-hoc adjustment procedures [@gelmanWhyWeUsually2012].

@@

```{r}
tmpr <- fits_wb %>% 
  filter(model > 1) %>% 
  mutate(
    sdx = map_dbl(fit, ~sd(.x$data[,4])),
    sdy = map_dbl(fit, ~sd(.x$data[,1]))
  ) %>% 
  mutate(
    x = map(
      fit, 
      ~as_draws_df(.x, variable = "b_[i|m]_cw$", regex = TRUE) %>% 
        rename(b = 1) %>% 
        mutate(r = b * (sd(.x$data[,4]) / sd(.x$data[,1])), .keep = "unused") %>% 
        describe_posterior(
          centrality = "mean", 
          ci = 0.95, 
          ci_method = "ETI", 
          test = "pd",
          diagnostic = NULL
        ) %>% 
        as_tibble())
  ) %>% 
  mutate(model = factor(model, labels = c("Internet", "Mobile"))) %>% 
  unnest(x) %>% 
  mutate(
    result = str_glue(
      "{number(Mean, .001)} [{number(CI_low, .001)}, {number(CI_high, .001)}]"
    ),
    .keep = "unused"
  ) %>% 
  select(outcome, model, result)
tmpr
```

In addition to these recommended raw associations [@baguleyStandardizedSimpleEffect2009], we also computed the standardized average associations ($\alpha_4(\frac{\sigma_x}{\sigma_y})$). In line with the unstandardized coefficients reported above, the standardized associations were small according to common guidelines: (life satisfaction & internet = `r tmpr$result[1]`, life satisfaction & mobile = `r tmpr$result[2]`, negative experiences & internet = `r tmpr$result[3]`, negative experiences & mobile = `r tmpr$result[4]`, positive experiences & internet = `r tmpr$result[5]`, positive experiences & mobile = `r tmpr$result[6]`).

@@

Our model also measured the extent to which internet technology adoption predicted well-being between countries. These between-country associations indicated, with high confidence, that countries with greater average levels of internet and mobile broadband adoption tended to report greater average levels of life satisfaction and positive experiences, and lower average levels of negative experiences (Figure \@ref(fig:fig-gwp-data)B, Between-country associations). However, this pattern is likely to reflect multiple other causes, such as socioeconomic factors and levels of inequality, impinging on both well-being and internet technology adoption. For example while wealthier nations are likely to have greater internet coverage, they also tend to have greater healthcare coverage, which in turn is likely to affect levels of well-being.

@@

```{r}
tmpr <- fits_mh %>% 
  filter(model > 1) %>% 
  mutate(
    sdx = map_dbl(fit, ~sd(.x$data[,4])),
    sdy = map_dbl(fit, ~sd(.x$data[,1]))
  ) %>% 
  mutate(
    x = map(
      fit, 
      ~as_draws_df(.x, variable = "b_[i|m]_cw$", regex = TRUE) %>% 
        rename(b = 1) %>% 
        mutate(r = b * (sd(.x$data[,4]) / sd(.x$data[,1])), .keep = "unused") %>% 
        describe_posterior(
          centrality = "mean", 
          ci = 0.95, 
          ci_method = "ETI", 
          test = "pd",
          diagnostic = NULL
        ) %>% 
        as_tibble())
  ) %>% 
  mutate(model = factor(model, labels = c("Internet", "Mobile"))) %>% 
  unnest(x) %>% 
  mutate(
    result = str_glue(
      "{number(Mean, .001)} [{number(CI_low, .001)}, {number(CI_high, .001)}]"
    ),
    .keep = "unused"
  ) %>% 
  select(outcome, model, result)
tmpr
```

Although numerically greater than the unstandardized coefficients reported above, the standardized associations were small according to common guidelines: (anxiety & internet = `r tmpr$result[1]`, anxiety & mobile = `r tmpr$result[2]`, depression & internet = `r tmpr$result[3]`, depression & mobile = `r tmpr$result[4]`, depression & internet = `r tmpr$result[5]`, depression & mobile = `r tmpr$result[6]`).

@@

Our results do not provide evidence supporting the view that the internet, and technologies enabled by it such as smartphones with internet access, are actively promoting or harming either well-being or mental health globally. Our results also suggest that age- and sex-based differences in associations linking internet technology adoption and well-being are small. However, we did find evidence suggesting that internet technology adoption is more negatively associated with young individuals' mental health than older groups' (whose associations were, in some cases, positive). Nevertheless, these associations, although sometimes credibly different from zero, appeared small in magnitude. Because of their apparently small magnitude and the fact that the data are not at the individual level, we highlight that these contrasts do not speak to whether the associations might be clinically relevant.

It is important to note that the data and theory required to address this question at the causal level are absent. Consequently, our analyses cannot account for potential confounders in the associations linking mental health and well-being to internet technology adoption. Our descriptions, therefore, are suggestive but are not intended to provide evidence for, or against, causal relations. Although the first to study this topic within a global perspective, future studies should investigate other potential differences based on demographic, economic, and cultural factors, which we could not address here. In addition, our results regarding mental health, while comprehensive, are necessarily and considerably less certain than those on well-being, due to the lack of diverse global data and our own reliance on GBD estimates in place of objective measurements of mental health. In addition, the data aggregation necessary to link internet adoption and well-being at the national level preclude investigating other important and relevant questions, such as comorbidities.
